---
title: "V-Dem Data Lab"
subtitle: "Exploring V-Dem and V-Party in R"
author: "[Your Names]"
date: today
format:
  html:
    toc: true
    code-copy: true
    embed-resources: true
  typst: default
execute:
  warning: false
  message: false
---

## Getting started

This file is a toolbox for your individual exploration. Run the examples,
then change the countries, years or indicators to follow your own interests.
The aim is to understand a measure, discover a pattern and develop a research
question. You do not need to work through every example.

For our ~45-minute session, spend about 35 minutes exploring and keep the final
10 minutes for discussion. Choose either the V-Dem or the V-Party example.

In RStudio, run a code chunk using its green arrow. In Positron, use the >Run Cell
button above each code cell. **Render** in RStudio and **Preview** in Positron produces a 
document from the whole file. Write your observations in the spaces below.

## Preparation

### Install packages once

Run this chunk manually in the beginning of the session. It is excluded from rendering
(`eval: false`). Installation requires an internet connection and may take
a few moments. See the [official package repository](https://github.com/vdeminstitute/vdemdata)
for installation guidance.

```{r}
#| label: installation
#| eval: false

install.packages(c("tidyverse", "remotes"))
remotes::install_github("vdeminstitute/vdemdata")
```

### Load packages and datasets

```{r}
#| label: setup
if (!require(pacman)) install.packages("pacman")
p_load(tidyverse, vdemdata)

vdem_data <- vdemdata::vdem
vparty_data <- vdemdata::vparty
codebook <- vdemdata::codebook #vdem codebook only, vparty codebook available: 

# Record the installed package version.
packageVersion("vdemdata")
```

V-Dem here uses **country-year** observations. V-Party uses
**country-party-date** observations: do not assume that each party is observed
every year. Check the coverage of your chosen sample before comparing it.
The datasets have different time coverage and release schedules.

## 1. Inspect the data and find a measure

### Basic tools

```{r}
#| label: inspect-data
# Number of rows and columns
dim(vdem_data)
dim(vparty_data)

# A small preview, rather than printing the entire dataset
vdem_data |> select(country_name, year, v2x_polyarchy) |> head()
vparty_data |> select(country_name, year, v2paenname, v2paid) |> head()

# Available years
range(vdem_data$year, na.rm = TRUE)
range(vparty_data$year, na.rm = TRUE)
```

For an interactive spreadsheet-style view, use `view()`.
These commands are optional and excluded from rendering.

```{r}
#| eval: false
view(codebook)
view(vdem_data)
view(vparty_data)
```

### Search the V-Dem codebook

```{r}
#| label: search-codebook
# Search indicator names by keyword; replace the search term.
democracy_variables <- vdemdata::find_var("democracy")
democracy_variables |> select(tag, name) |> head(10)

# Ask about a specific variable using its name in quotation marks.
vdemdata::var_info("v2x_polyarchy")
```

`find_var()` returns a table of codebook entries, not just variable names.
`var_info()` expects a variable name, not a dataset. These helpers are
documented for the V-Dem codebook. For V-Party, consult its separate
[codebook on the dataset page](https://www.v-dem.net/data/v-party-dataset/).

You can search column names in either dataset. This finds names, not definitions:

```{r}
#| label: search-columns
names(vdem_data)[str_detect(names(vdem_data), "^v2x_")] |> head(15)
names(vparty_data)[str_detect(names(vparty_data), "vote|seat|corrupt")]
```

**Your notes:** Which variable interests you? What does it measure? What do
high and low values mean? What aspect of your interest does it leave out?

*Write here.*

## 2A. V-Dem: Compare countries over time

### Select countries, years and columns

```{r}
#| label: select-vdem
# Change these settings to explore your own cases.
chosen_countries <- c("Germany", "Hungary", "Poland")
first_year <- 1990
last_year <- 2024

democracy_sample <- vdem_data |>
  filter(country_name %in% chosen_countries,
         year >= first_year, year <= last_year) |>
  select(country_name, year, v2x_polyarchy, v2x_libdem)

head(democracy_sample)

# Check how many observations and missing values each country has.
democracy_sample |>
  group_by(country_name) |>
  summarise(observations = n(),
            missing_electoral = sum(is.na(v2x_polyarchy)))
```

`filter()` selects rows; `select()` selects columns. `%in%` matches any of the
countries in your list. `|>` passes the result to the next step. `NA` means
missing data, not zero.

### Draw a line graph

```{r}
#| label: plot-vdem
democracy_plot <- ggplot(democracy_sample,
                        aes(x = year, y = v2x_polyarchy,
                            colour = country_name)) +
  geom_line(linewidth = 0.8) +
  labs(title = "Electoral democracy over time",
       x = "Year", y = "Electoral Democracy Index",
       colour = "Country") +
  theme_minimal()

democracy_plot

var_info("v2x_polyarchy")
var_info("v2x_libdem")
```

**Try it:** Replace `v2x_polyarchy` in the plot with `v2x_libdem`, and adjust
the title and axis label. Consult `var_info("v2x_libdem")` first. These indices
capture different concepts; a difference between their values is not itself
a measure of institutional change.

**Your notes:** Describe one pattern. Does it match your expectations?
Separate the observation from your possible explanation.

*Write here.*

## 2B. V-Party: Compare parties at one election date

Party names are stored in `v2paenname` (English name), and party identifiers
in `v2paid`. The example uses vote shares, `v2pavote`; consult the V-Party
codebook for its definition, scale and treatment of electoral alliances.

```{r}
#| label: select-vparty
chosen_country <- "Germany"

# Inspect available dates before choosing one.
vparty_data |>
  filter(country_name == chosen_country) |>
  distinct(year, historical_date) |>
  arrange(desc(year), historical_date) |>
  head(10)

# Example: German parties observed in 2017.
party_sample <- vparty_data |>
  filter(country_name == chosen_country, year == 2017) |>
  select(country_name, year, historical_date,
         v2paid, v2paenname, v2pavote)

party_sample
```

Check that your sample refers to one election date before making the chart.
If your chosen year contains several dates, add a filter for `historical_date`.
Missing parties or missing vote shares should not be interpreted as zero votes.

```{r}
#| label: plot-vparty
party_sample |> summarise(missing_vote_shares = sum(is.na(v2pavote)))

party_plot <- ggplot(party_sample |> filter(!is.na(v2pavote)),
       aes(x = reorder(v2paenname, v2pavote), y = v2pavote)) +
        geom_col(fill = "steelblue") +
        coord_flip() +
        labs(title = "Recorded party vote shares: Germany, 2017",
            x = NULL, y = "Vote share (%)") +
        theme_minimal() +
        theme(
            plot.title.position = "plot",
            plot.title = element_text(hjust = 0.5)
        )

party_plot

```

**Try it:** Choose another country and an available election year. Update the
chart title. Alternatively, find a party characteristic in the V-Party
codebook, locate its column and explore it. Use party identifiers when
tracking parties; names alone may not reliably identify them over time.

**Your notes:** What does your comparison show? What would you need to know
about the election or party coverage to interpret it?

*Write here.*

## 3. Develop a research question

Based on your exploration, write down:

1. One research question.
2. The countries, parties and period you would investigate.
3. One additional variable or data source you would need.
4. Something your current graph cannot establish.

A descriptive pattern can inspire an explanation, but does not establish
causality. Where uncertainty intervals are provided, consider them before
interpreting small differences.

*Write here.*

## Optional tools: Save your results

Run manually if you want to save your V-Dem sample and plot in the folder
containing this file. These commands are excluded from rendering.

```{r}
#| eval: false
write_csv(democracy_sample, "my-vdem-sample.csv")

ggsave("my-democracy-plot.png", plot = democracy_plot,
       width = 8, height = 5, dpi = 300)

ggsave("my-party-plot.png", plot = party_plot,
       width = 8, height = 5, dpi = 300)
```

## Sources and reproducibility

- [vdemdata package documentation](https://github.com/vdeminstitute/vdemdata)
- [V-Dem data and codebook](https://www.v-dem.net/data/the-v-dem-dataset/)
- [V-Party data and codebook](https://www.v-dem.net/data/v-party-dataset/)

Record the package and dataset versions you use and cite the datasets in
research. Values may be revised between releases: use one release throughout
your comparison. HTML is the default output; select Typst when you want a PDF.

```{r}
#| label: session-information
#| include: false
sessionInfo()
```